P3064R2: How to Avoid OOTA Without Really Trying

### 2024-08-31 paper/ootavacuous.tex
ootavacuous: Convert two by-hand listings to tables
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = y.load(memory_order_relaxed);
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple OOTA}
\label{lst:Simple OOTA}
\end{listing}

Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
shows a simple example where an OOTA cycle might result in all of \co{x},
\co{y}, \co{r1}, and \co{r2} having final values of 42, despite the fact
that there is nothing in the initial values or the executable code to
support such an outcome:

\begin{enumerate}
\item   Line~6 loads from \co{x} into \co{r1}, claiming to read
	the value of line~13's store rather than \co{x}'s initial value
	and somehow obtaining 42.
\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
\item   Line~12 loads 42 from \co{y} to \co{r2}.
\item   Line~13 stores 42 to \co{x}, justifying the value loaded by line~6.
\end{enumerate}

@@

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = y.load(memory_order_relaxed);
13   x.store(42, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple Reordering}
\label{lst:Simple Reordering}
\end{listing}

It is important to distinguish true OOTA cycles from OOTA-like
behavior caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder
lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 as follows:
\begin{enumerate}
\item   Line~13 stores 42 to \co{x}.
\item   Line~6 loads 42 from \co{x} into \co{r1}.
\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
\item   Line~12 loads 42 from \co{y} to \co{r2}.
\end{enumerate}
Current C++ implementations can and do exhibit this reordering behavior.

>>>>

\begin{listing}[tbp]
% Constructed by hand.
{
\scriptsize
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 =\textsubscript{rlx} x} &
	\texttt{~~int r2 =\textsubscript{rlx} y} \\
2. &
\texttt{~~y =\textsubscript{rlx} r1} &
	\texttt{~~x =\textsubscript{rlx} r2} \\
\end{tabular}

\vspace{0.1in}
Condition (0:r1=42 $\wedge$ 1:r2=42)
sometimes satisfied.
}
\caption{Simple OOTA}
\label{lst:Simple OOTA}
\end{listing}

Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
shows a simple example where an OOTA cycle might result in all of \co{x},
\co{y}, \co{r1}, and \co{r2} having final values of 42, despite the fact
that there is nothing in the initial values or the executable code to
support such an outcome.
First, P0's line~1 loads from \co{x} into \co{r1}, claiming to read the value
of P1's line-2 store rather than \co{x}'s initial value and somehow
obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
loaded by P0's line~1.

@@

\begin{listing}[tbp]
% Constructed by hand.
{
\scriptsize
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 =\textsubscript{rlx} x} &
	\texttt{~~int r2 =\textsubscript{rlx} y} \\
2. &
\texttt{~~y =\textsubscript{rlx} r1} &
	\texttt{~~x =\textsubscript{rlx} 42} \\
\end{tabular}

\vspace{0.1in}
Condition (0:r1=42 $\wedge$ 1:r2=42)
sometimes satisfied.
}
\caption{Simple Reordering}
\label{lst:Simple Reordering}
\end{listing}

It is important to distinguish true OOTA cycles from OOTA-like
behavior caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder
lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 via the following steps.
First, P1's line~2 stores 42 to \co{x}.
Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
Current C++ implementations can and do exhibit this reordering behavior.
